```{r}

```

```{r}
#Qtn 1
#set seed 1
set.seed(1)


#sample 100 obs
expData <- rexp(1000, 2)
max(expData)
```

```{r}
hist(expData, prob = TRUE)
curve(dexp(x, rate = 2), col='blue', add = TRUE)
curve(dexp(x, rate = 1.940473), col='red', add = TRUE)

```



```{r}
logl <- function(lambda, xData){
  loglpr <- length(xData)*log(lambda) - lambda*sum(xData) 
  return(-loglpr)
}
```

```{r}
optim(par = 5, 
      fn = logl, 
      xData= x, 
      method = 'Brent', lower = 0, upper = 5,
      )#control=list(fnscale=-1) #by default optim minimises, the fn scale is to maximiz
```

```{r}
opt <- optimise(f = logl, xData = xData, lower=0.001, upper=10)
plot.new()
hist(xData, main="", xlab="x", prob=TRUE)

xr <- seq(from=0.01, to=3, length.out=500)

#take note how the estimated parameter is extracted from the list.
yr.est <- dexp(xr, rate = opt$minimum)

lines(xr, yr.est, col="blue")
yr.true <- dexp(xr, rate = 2)
lines(xr, yr.true, col="red")
legend("topright", legend = c("Fitted", "True"), col=c("blue", "red"),
lty=c(1,1), bty="n")
```

```{r}
normData <- rnorm(10000, mean = 0, sd = 1)

loglike <- function(param, data){
  n <- length(normData)
  mu <- param[1]
  s2 <- param[2]**2
  
  loglikepr <- -n*0.5 * log(s2) - 0.5*sum( (normData-mu)**2 )/s2
  return(-loglikepr)
}
```

```{r}
opt <- optim(c(5,5), fn=loglike,
    method="L-BFGS-B",
    lower=c(-Inf, 0.001),
    upper=c(Inf, Inf),
    data = normData,
    )#control = list(fnscale = -1)

opt


#the mean and sd estimates
cat("\n Parameter estimate = ", opt$par)
##
## Parameter estimate = 4.92677 2.856681
#the value is 0 and thus the algorithm converged
cat("\n Convergence message = ", opt$convergence)
##
## Convergence message = 0
#check the solutions
cat("\n mean = ", mean(normData))
```

```{r}

#Contor plots
#range of mu and s.star
xr.mu <- seq(from=1, to=8, length.out=100)
xr.s.star <- seq(from=0.5, to=2, length.out=100)

logl.c <- function(mu, s.star, xData){
#return the loglikelihood function for Gaussian data
#par = c(mu, s*); mu = mean(X), s = log(sd(X))
#transform from sigma* to sigma
s <- exp(s.star)
#the variance of X
s2 <- s**2
n <- length(xData)
temp <- 0
for (i in 1:n){ temp <- temp + (xData[i]-mu)**2 }
logl <- -n*0.5*log(s2) - 0.5*temp/s2
return( logl[1] )
}
```


```{r}
logl.vec <- Vectorize(FUN=logl.c, vectorize.args = c("mu", "s.star"))
outs <- outer(X=xr.mu, Y=xr.s.star, FUN=logl.vec, xData=xData)
#produce the contour plot
contour(xr.mu, exp(xr.s.star), outs,
nlevel=30, #the number of levels of the plot
xlab=expression(mu),
ylab=expression(sigma))
```

#outer function

The outer function requires two inputs X and Y and applies these
values to your function.
X and Y should be vectors such that all element pairs are evaluated
by the outer function.
Previously we saw that the first row of the output is 5, 6 and 7.
This is because the first row of the output matrix is 1 × 5, 1 × 6 and
1 × 7. The third row similarly is 3 × 5, 3 × 6 and 3 × 7.
The outer function evaluates much faster than using a nested loop
to produce the above output and is preferred.

```{r}
x1 <- 1:3
x2 <- 5:7
func1 <- function(x,y){
x*y
}
outer(X=x1, Y=x2, FUN=func1)
```

```{r}

```

