---
title: 'Multivariate Analysis: Continuous Assessment 2'
author: "Molly Ryan (RYNERI002) Tinotenda Mutsemi (MTSTIN007)"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r, echo=F}
#set the working directory
#setwd('C:\\Users\\mutse\\OneDrive\\Desktop\\UCT\\DataScience1\\Multivariate Analysis\\Scripts\\CA2')
```

```{r, echo=F, results='hide', message=FALSE, warning=FALSE}
#load the libraries
invisible(library(corrplot))
library(psych)
```

Singular Value Decomposition was used to find lower rank approximations of X, ranging from rank 12 (full rank) down to rank 1. Then the error for each approximation of X was calculated.


```{r, echo=F, results = F}
#read txt file
X <- read.table("CA2.txt", header = TRUE)
# head(X)
#make X a matrix
X <- as.matrix(X)
head(X)
```


```{r, echo=F}
#For each approximation Xk of rank k, 

XtrX <- t(X) %*% X

#calculate the SVD of XtrX
svdX <- svd(X)

#find the best rank k approximation for k = 1 to num of columns in X
m <- dim(X)[2]
#create a list of 12 b matrices
B <- list()


B[[1]] <- prod(svdX$d[1]) * (svdX$u[, 1] %*% t(svdX$v[, 1]))
colnames(B[[1]]) <- colnames(X)
for (k in 2:m) {
  b<-svdX$u[,1:k] %*% diag(svdX$d[1:k]) %*% t(svdX$v[,1:k])
  B[[k]] <- b
  colnames(B[[k]]) <- colnames(X)
}

```


```{r, echo=F}
#calculate the error for each rank k approximation
error <- list()
for (k in 1:length(B)) {
  error_k <- X - B[[k]]
  error[[k]] <- error_k
}
```

The mean vector of the approximation error for the rank 3 approximation can be seen below.


```{r, echo=F, results='asis'}
#Qtn 1: we find mean error vector error3
error3 <- error[[3]]
#mean for each column
mean_error3 <- colMeans(error3)
cat("$$ Δ_3 = \\begin{pmatrix}", paste(round(as.vector(mean_error3),2), collapse = " \\\\ "), "\\end{pmatrix} $$", sep = " ")

```


```{r, echo=F, results='hide'}
#Qtn 2
std_error3 <- scale(error3)
```


A histogram plotting the standardised error for the rank 3 approximation matrix follows. It can be seen that the error is approximately normally distributed.

```{r, echo=F, fig.align='center'}
hist(std_error3, main="Histogram showing the standardised error of the rank 3 approximation", xlab="Standardised Error")
```




When comparing the correlations between the variables in X to the correlations between the variables in the rank 3 approximation, a clear difference can be seen. This is clear in the heatmaps plotted below. Indeed, it appears that the lower the rank of the approximation, the less correlation is preserved.



```{r, echo=F, fig.align='center'}
par(mfrow=c(1,2))
suppressWarnings({
  corrplot(cor(X), t1.col = "black", title="Correlation heat map for \n the full   matrix X", mar=c(1,1,5,1))
  corrplot(cor(B[[3]]), t1.col = "black", title="Correlation heat map for \n the rank 3 approximation matrix", mar=c(1,1,5,1))
})
#the lower the rank, the less the correlation is preserved
```

```{r, echo=F}

frob_vec <- c()
for (i in 1: 12){
  frob <- norm(error[[i]], type = "F")
  # frob <- sqrt(tr(error[[i]]%*%t(error[[i]])))[1]
  frob_vec <- rbind(frob_vec, frob)
}
#covert to numeric
frob_vec <- as.numeric(frob_vec)

```

Lastly, the Frobenius Norm of the matrix can be seen to decrease with increasing rank (k). This relationship is plotted below.

```{r, echo=F, fig.align='center'}
frob_vec <- cbind(frob_vec, seq(1,12))
plot(frob_vec[,2], frob_vec[,1], type = "l", xlab = "Rank", ylab = "Frobenius Norm")

```

```{r, echo=F}
```



